A general and efficient representation of Ancestral Recombination Graphs

### 2023-10-06 paper.tex
Implement most of Gideon and Alex's comments

\section*{Introduction}
Estimating the genetic genealogy of a set of genome sequences
under the influence of recombination,
usually known as an Ancestral Recombination Graph (ARG), is a long-standing
goal in genetics.
% TODO say this better.
From a high-level, an ARG describes the different paths of genetic inheritance
caused by recombination, and encodes a sequence of correlated genealogical
trees along the genome.
Recent breakthroughs
in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
schaefer2021ancestral,wohns2022unified,zhang2023biobank,zhan2023towards}
and data representation~\citep{kelleher2016efficient,kelleher2018efficient}
have raised the realistic prospect of ARG-based analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using inferred ARGs as input have begun to
appear~\citep{osmond2021estimating,fan2022genealogical,hejase2022deep,zhang2023biobank,
nowbandegani2023extremely}
and many more are sure to
follow~\citep{harris2019database,harris2023using}.

% TODO need to mention ARG-as-a-data-structure to clarify.
% What is the problem?
Although it is widely accepted that ARGs are important, there is significant
confusion about what, precisely, an ARG \emph{is}.
Originally, ARGs were defined by Griffiths and colleagues as an alternative
formulation of the coalescent with recombination~\citep{hudson1983properties},
where the stochastic process of coalescence and recombination
among ancestral lineages is formalised as a
% Note: please don't add "mathematical graph". If a reader doesn't know what a
% graph is they're not going to understand this paper.
graph~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
Subsequently, ARGs have come to be thought of as a data
structure~\citep{minichiello2006mapping}, i.e.\ describing
the outcome of such a random process or an inference
of the genetic genealogy of a sample of genome sequences.
However, the distinction between process and data structure is not clear cut
and subfields use the term
differently (see Appendix XXX).

The word ``ARG'' is now widely used to refer to genetic genealogies in
general~\citep[MORE][]{mathieson2020ancestry}, including
varied outputs from modern simulation and
inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
kelleher2019inferring, speidel2019method, baumdicker2021efficient, zhang2023biobank}.
This broad usage means there is disagreement on the
exact object being defined. This causes problems, for example,
when evaluating inferred ARGs, and even when defining
what the criteria for success should be.
It also means there is no shared format for interchanging ARG data between
programs, resulting in the assorted negative consequences with which
geneticists are only too familiar~\citep{excoffier2006computer}.
Perhaps most importantly, the direct linkage to
the original stochastic process definition leads to an
overly narrow interpretation of ARGs, which is significantly
out of step with modern methods and large-scale datasets, and
can lead to missed opportunities in their application and usage.
Such problems go much deeper than pedantic discussion over terminology;
they directly inform what we should be attempting to estimate, how we
assess success, and the expected limitations on the precision
and accuracy of our inferences.

@@

There are some important consequences to these definitions.
Firstly, there is
no limitation on mating system, ploidy, or age structure in a gARG.
By making the
basic unit of analysis an individual's genomes,
any pattern of inheritance among mixed ploidy (e.g.\ haplodiploid
species) can be accommodated.
Secondly, because we record the intervals of inheritance
between genomes,
arbitrary patterns of genetic inheritance such as
gene conversions,
multiple simultaneous recombinations
(e.g.\ Fig.~\ref{fig-arg-in-pedigree}),
and so on can be directly expressed.
% Wilder: give forward ref to section on efficiency here, mention tree sequence
Thirdly, local trees that we generate for a particular position on the
genome can have nodes of any arity, including ``unary''
nodes with a single child (see Section XXX).
% TODO should rephrase this from parent/child to ancestor/descendant?
Fourth, recombination is modelled in terms of
its outcomes in terms of inheritance intervals among parental genomes;
thus we will have a child node (the recombinant; e.g.\ node \noderef{a}
Fig.~\ref{fig-arg-in-pedigree}) and two (or more; see section XXX)
paternal genomes from which this child has inherited segments.
% TODO rephrase the above, it'll confuse people. Just need to reiterate the
% points about events, and maybe lead into the next section
Finally, gARGs are not defined in terms of the
coalescent with recombination (or any other) stochastic process.
Thus, the gARG encoding is focused on faithfully representing the
\emph{outcome} of the evolutionary processes of genetic inheritance
with recombination,
and is agnostic to the details of those processes.
% The paragraph above is a bit weak and muddled, needs a few more passes.
% In particular, the "Finally" bit doesn't really seem to follow
% from the others. Is there a more direct way we can address these points ?

@@

\section*{Event ARGs}\label{eARG}
Genome ARGs are composed by
describing the details of genetic inheritance
between specific ancestor and descendant genomes.
In contrast, ARGs are classically seem defined
by the evolutionary \emph{events} in the history of a sample.
The difference between gARGs and this prevalent view of
what defines an ARG is subtle and the distinctions sometimes porous.
It is therefore useful to first survey some representative
definitions of
an ARG from the literature, to allow us to capture the common features
of these definitions
(see Appendix XXX for a full review).

\citet{brandt2021evaluation} follow the original Griffiths definition of
an ARG by coupling it directly to a generative process, stating
that the ``full ancestral recombination graph (ARG) is a structure that encodes all
coalescence and recombination events resulting from the stochastic process of
the coalescent with recombination.''
\citet{shipilina2023origin} emphasise the true existance of the
corresponding events (rather than the generative process) by saying that
``[t]he ARG describes the complete
ancestry of a sample of genomes through a series of real coalescence and
recombination events.''
Following this trend to decouple the concept of an ARG from
generative processes~\citep{minichiello2006mapping},
\cite{mathieson2020ancestry}
define an ARG as ``a subset of the pedigree...
contain[ing] only those edges along which inherited segments of DNA have been
transmitted, and only those nodes corresponding to ancestors in which there was
a recombination or coalescence event.''
In contrast, ~\citet{gusfield2014recombinatorics} defines an ARG as [TODO]
x
% Papers that do not discuss how positions work:
% - Mathieson and Scally: no discussion of how it works concretely (but quite
%   close to gARG, in some ways)
% TODO list out some of these as examples of where people don't bother
% getting into this important detail?
The common feature of these definitions is that they are expressed
in terms of evolutionary \emph{events}, and so for the purposes of
discussion we can label
this prevailing interpretation as an ``Event ARG'', or eARG.
(Many definitions also emphasise
the necessity of exhaustively capturing \emph{all} events, which we
examine in later sections.)
Another common feature of these definitions is that they do not explicitly
discuss the mechanism by which genealogical trees vary along the genome.
The topology of an eARG defines the common ancestor and
recombination events that occured,
but the graph structure alone is not sufficient to define the local trees,
or equivalently, how ``ancestral material'' (Section XXX) flows along
the edges of the ARG.
This is usually assumed to follow the mechanism described by
Griffiths and colleagues in which we have
two different ``types'' of node in the graph:
common ancestor nodes in which the inbound lineages are merged into a
single ancestral lineage with one parent, and recombination
nodes in which a single lineage is split into two independent
ancestral lineages.
Recombination nodes are then
annotated with
the corresponding crossover breakpoint~\citep{griffiths1996ancestral}.
Fig.~\ref{fig-arg-data-structure} shows an example of a classical
eARG with three samples (\noderef{a}, \noderef{b}, and \noderef{c})
and a single recombination event, along with an example concrete
encoding of the data.  There is a single recombination event
at node \noderef{d} with a breakpoint at position $x$. We
assume that \noderef{d} inherits genetic material to the
left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.
As shown in Fig.~\ref{fig-arg-data-structure}B,
the local trees are embedded in the graph, and can be reconstructed
in essentially the same manner as outlined the previous section for
gARGs. The only difference is that decisions about which outbound
edge to follow during the rootward traversal are taken at recombination
nodes, following
either the common convention that the order of parents
is significant~\citep[e.g.][]{griffiths1991two},
or by incorporating additional information to
distinguish left and right parents
into the
encoding~\citep[e.g.][]{gusfield2014recombinatorics,ignatieva2021kwarg}.

@@

\section*{Prospective vs retrospective ARGs}
Event ARGs, as defined in the previous section as a way of encompassing
the range of definitions provided in the literature, are
inherently \emph{retrospective}.
Event ARGs are defined in terms
of the events that occur in the history of some set of sampled
sequences.
For example, Fig.~\ref{fig-arg-in-pedigree}
is retrospective because it only includes information
about genetic inheritance relative to the sample individuals
$D_1$ and $D_2$ (genomes \textsf{a}--\textsf{d}).
% Gregor & Jerome think that "inheritance" is clearer here than "transmission"
Genetic inheritance must have occurred between (e.g.)
$D_8$ and her daughter $D_6$, but this is not recorded because
% this is probably not true, given we know the complete pedigree?
it is not relevant to the genetic ancestry of the samples.

@@

Prospective ARGs arise in individual-based forwards-time
simulations~\citep{kelleher2018efficient,haller2018tree},
where the genetic inheritance between genomes is recorded
exhaustively and periodically
``simplified'' to retain only
the ancestry relevant to the current population.
\citet{kelleher2018efficient} refer to the structure
recording all such forwards-in-time genetic relationships
as an ``embellished pedigree`` (and suggest the term ``nedigree'' as a shorthand).
Describing this instead as a ``prospective ARG'', emphasising
the use of the same data structures and the direct connection with
the usual retrospective interpretation of an ARG, may be
more useful.
The ``simplify'' algorithm, which prunes away parts of the graph that
are not ancestral to a sample, is the key enabling factor for
this approach to forwards-time simulation and is
described in detail by
\citet{kelleher2018efficient}. It is a
powerful tool with many other applications,
some of which we discuss in the following sections.

\section*{Converting an eARG to a gARG}
Any eARG can be converted to a gARG without loss of information,
and the conversion process is very similar to
simplifying a prospective gARG with respect to a set of samples.
The process involves two steps, and is illustrated in
Fig.~\ref{fig-ancestry-resolution}.
The first step is straightforward
(Fig.~\ref{fig-ancestry-resolution}B).
We duplicate the nodes and
edges from the input eARG, and the add inheritance annotations
to the output gARG's edges. If the node is a common ancestor node,
we annotate the single outbound edge with the interval $[0,L)$,
for a genome of length $L$.
If the node is a recombination event with a breakpoint $x$,
we annotate the outbound edges with the intervals $[0, x)$
and $[x, L)$.
These inheritance annotations are clearly in one-to-one
correspondance with the information in the input eARG,
but they are a superset of the
``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
of the sample.

@@

\section*{ARGs and local trees}
Local trees, describing the genetic genealogy of a sample
at a particular location along the genome, are said to be
``embedded'' within in ARG [CITES], and the relationship
between the two is subtle and important.
% An ARG is sometimes referred to as being equivalent to a set of
% local trees along the genome [TODO citation], but the connection
% between the two is subtle.
% In order for the mapping to be one-to-one, so that the original
% gARG can be exactly recovered from a set of local trees and
% their genome coordinates,
% some specific properties of the ARG and the trees are required.
Note that many authors use the term ``marginal''
trees~\citep[e.g.][]{griffiths1996ancestral,minichiello2006mapping} [more]
which we avoid due the potential confusion with the
statistical concept of a marginal distributions indicating
an integrated tree over many positions.

@@

The nodes in a local tree encoded
in this way are \emph{labelled} (i.e.\ carry the identity
of the corresponding ARG node), and this is fundamentally
important to how a local tree relates to its parent ARG.
If we are given the sequence of local trees for a gARG
encoded as an oriented tree
and along with the genome interval covered by each tree,
then we can recover precisely the same ARG. Thus, there
is a one-to-one correspondance between an ARG and
the sequence of local trees that it encodes.

The internal nodes of trees, however, are usually
seen as \emph{unlabelled}, since they most often correspond
to anonymous hypothetical ancestors. Internal
ARG nodes likewise correspond to anonymous hypothetical
ancestors, but if we do not label them we lose the fact that
the \emph{same} ancestors persist across multiple trees.
Without labelled internal tree nodes,
we must infer their identity
from the tree topology, which is both
expensive to compute and imprecise. For example
we could infer that node \textsf{k} is shared in the second
and third trees of in Fig.~\ref{fig-simplification}D
because it subtends identical subtrees in both,
but it would not be straightforward to infer that it
is also the root of the first tree.

\cite{shipilina2023origin} discuss the same ideas, and note that
``full ARG ... contains more information than the series of tree
sequences along the genome''. They view the ``full ARG'' as an
eARG, and the ``series of tree sequences'' is their
% jk: probably should delete this. It *is* muddled though
(somewhat muddled)
term for the sequence of local trees along the genome.
Their statement is true, \emph{if} the local trees are
unlabelled and contain
no unary nodes. However, as we have shown, if the local trees
contain labelled internal (possibly unary) nodes,
then  there is precisely as
much information in them as an ARG. Whether such unary nodes
are \emph{observable} is an entirely separate question,
of course.

These properties have interesting consequences when we
consider the inference strategy of estimating local trees
independently and subsequently stitching them together
into an ARG.
Two recent methods take this approach,
\relate\ and
\espalier~\citep{rasmussen2022espalier},
and have different strategies for reconciling the local trees into an ARG.
\relate\ first infers local tree topologies, and the identifies
edges that persist from tree to tree by reasoning about the sets of samples
subtended. It does not attempt to be exhaustive.
\espalier, on the other hand, is specifically interested
in the details of recombination events (among small samples)
and therefore attempts to infer
the precise subtree prune-and-regraft (SPR)
operations~\citep{hein1990reconstructing,song2003on,song2006properties}
induced by recombination.
Inferring the SPRs between leaf-labelled trees is
NP-hard~\citep{hein1996complexity,allen2001subtree,bordewich2005computational},
but it is unclear what the complexity is when there
is a degree of internal node sharing between trees.

@@

\section*{Levels of simplification}\label{ARG_simplification}
ARG simplification is a powerful tool, and can do much more
besides removing unreachable nodes and edges in a prospective ARG
and resolving ancestral material in an eARG.
In general, we can think of
ARG simplification as the process
of removing nodes and re-writing edges (and their inheritance annotations)
to remove various types of redundancy.
The redundancy that we are interested in
revolves around the presence of nodes which appear
``unary'' in one or more local trees
(where a node has exactly one child over a particular genomic region).
These arise in several different ways.
Nodes with one child are not a standard feature of evolutionary trees;
we are usually interested in
nodes that have at least two children, because these
are in principle detectable from mutational differences
arising in the respective subtrees.
A tree branch may correspond to many generations
of individuals (or indeed cell divisions) and there is usually
very little information about these intermediates, or utility
in modelling them.
% FIXME the bridge between the next few paragraphs is a bit weird, in
% that we've just been talking about how interesting unary nodes are
% but then go on about how to remove them. Just needs a bit of smoothing
% out I think.
ARGs, however, are different, and there
can be significant information about evolutionary processes
provided by nodes that are unary in one or more local trees.
[FIXME this is clumsy; should mention somewhere, though.]
\citet{mathieson2020ancestry} show examples in which the ARGs
contain unary nodes, and state that paths are ``implicitly passing through points representing specific
individuals.''
% Full quote from fig caption:
% Genetic ancestry in the form of the ARG for a single individual. Combining
% genetic ancestry from different positions leads to a graph, incorporating all
% realized genetic ancestry paths, implicitly passing through points representing
% specific individuals.

@@

Simplifying away diamonds will remove many unary nodes from the
local trees, but there can still be nodes that are unary in all
of the local trees. In particular, a node can represent a recombinant
with multiple parents in the graph but only a single child (e.g.\ node \noderef{n}
in Fig.~\ref{fig-simplification}B), or can represent a common ancestor with
multiple children in the graph but in which no coalescence takes place
in the local trees
(node \noderef{r} in Fig.~\ref{fig-simplification}B).
% point out that this is why Hudson refers to CA nodes, not coalescence nodes
Such nodes are not singly connected in the graph, but are nevertheless unary in
all of the local trees.
The operation to remove them
therefore requires knowledge not just of the graph topology but also of the
ancestral material associated with the edges.
As we see in Fig.~\ref{fig-simplification}C,
removal of recombinant nodes can produce graph nodes with
more than two parents (e.g.~node \noderef{e}); and likewise, removal of
common ancestor but non-coalescent nodes can produce graph nodes with
more than two children (e.g.~node \noderef{s}). These represent a ``stacking up'' of
multiple---often indistinguishable---events into a single node.

@@

The remaining nodes are MRCAs of some subset of the samples
at \emph{some} positions along the genome. We still have
some unary nodes in the local trees, but these nodes will
correspond to a coalescence in at least one other
local tree. For example, node  \noderef{k} is unary in the fourth tree
of Fig.~\ref{fig-simplification}C, but is either binary
or ternary in all subsequent trees (recall this is a Wright-Fisher
simulation). The final level of simplification is to alter the edge annotations
such that, although no nodes are removed from the graph, all
unary nodes disappear from the local trees (Fig.~\ref{fig-simplification}D).
Note that although this last stage produces simpler local trees, by
removing information about the exact paths taken by lineages through
the graph, we lose potentially useful information about shared edges
between trees. The extent to which this information can be inferred
and utilised is an interesting open question.
% I don't think there's an actual manuscript or title to refer to here
% [CITE Ralph in prep.].

>>>>

\section*{Introduction}
Estimating the genetic genealogy of a set of genome sequences
under the influence of recombination,
usually known as an Ancestral Recombination Graph (ARG), is a long-standing
goal in genetics.
% TODO say this better.
From a high level, an ARG describes the different paths of genetic inheritance
caused by recombination, and encodes a sequence of correlated genealogical
trees along the genome.
Recent breakthroughs
in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
schaefer2021ancestral,wohns2022unified,zhang2023biobank,zhan2023towards}
and data representation~\citep{kelleher2016efficient,kelleher2018efficient}
have raised the realistic prospect of ARG-based analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using inferred ARGs as input have begun to
appear~\citep{osmond2021estimating,fan2022genealogical,hejase2022deep,zhang2023biobank,
nowbandegani2023extremely}
and many more are sure to
follow~\citep{harris2019database,harris2023using}.

% TODO need to mention ARG-as-a-data-structure to clarify.
% What is the problem?
Although it is widely accepted that ARGs are important, there is significant
confusion about what, precisely, an ARG \emph{is}.
Originally, ARGs were defined by Griffiths and colleagues as an alternative
formulation of the coalescent with recombination~\citep{hudson1983properties},
where the stochastic process of coalescence and recombination
among ancestral lineages is formalised as a
% Note: please don't add "mathematical graph". If a reader doesn't know what a
% graph is they're not going to understand this paper.
graph~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
Subsequently, an ARG  has come to be thought of as a data
structure~\citep{minichiello2006mapping}, i.e.\ describing
the outcome of such a random process or an inference
of the genetic genealogy of a sample of genome sequences.
However, the distinction between process and data structure is not clear cut
and subfields use the term
differently (see Appendix XXX).

The word ``ARG'' is now widely used to refer to genetic genealogies in
general~\citep[MORE][]{mathieson2020ancestry}, including
varied outputs from modern simulation and
inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
kelleher2019inferring, speidel2019method, baumdicker2021efficient, zhang2023biobank}.
This broad usage means there is disagreement on the
exact object being defined. This causes problems, for example,
when evaluating inferred ARGs, and even when defining
what the criteria for success should be.
It also means there is no shared format for interchanging ARG data between
programs, resulting in the assorted negative consequences with which
geneticists are only too familiar~\citep{excoffier2006computer}.
Perhaps most importantly, the direct linkage to
the original stochastic process definition leads to an
overly narrow interpretation of ARGs, which is significantly
out of step with modern methods and large-scale datasets, and
can lead to missed opportunities in their application and usage.
Such problems go much deeper than pedantic discussion over terminology;
they directly inform what we should be attempting to estimate, how we
assess success, and the expected limitations on the precision
and accuracy of our inferences.
% Thanks Gideon and Alex!
If not addressed, these problems will
impede progress on applying genetic genealogical
information to understand the biological world.

% Evidence of confusion to cite above in "problem" para?
% https://github.com/tskit-dev/what-is-an-arg-paper/issues/334
% The output of these programs is certainly not a collection of
% \emph{unrelated} local trees, as implied by,
% e.g.,
% \citet{hejase2020summary} when they state that \tsinfer\
% ``does not explicitly infer an ARG but rather a sequence of local
% gene trees''.
% More cites? Not worth harping on about now?
% https://github.com/tskit-dev/what-is-an-arg-paper/issues/38

@@

There are some important consequences to these definitions.
Firstly, there is
no limitation on mating system, ploidy, or age structure in a gARG.
By making the
basic unit of analysis an individual's genomes,
any pattern of inheritance among mixed ploidy (e.g.\ haplodiploid
species) can be accommodated.
Secondly, because we record the intervals of inheritance
between genomes,
arbitrary patterns of genetic inheritance such as
gene conversions,
multiple simultaneous recombinations
and so on can be directly expressed.
% Wilder: give forward ref to section on efficiency here, mention tree sequence
Thirdly, local trees that we generate for a particular position on the
genome can have nodes of any arity (number of child nodes), including ``unary''
nodes with a single child (see Section XXX).
% TODO should rephrase this from parent/child to ancestor/descendant?
Fourth, recombination is modelled in terms of
its outcomes in terms of inheritance intervals among parental genomes;
thus we will have a child node (the recombinant; e.g.\ node \noderef{a}
Fig.~\ref{fig-arg-in-pedigree}) and two (or more; see section XXX)
paternal genomes from which this child has inherited segments.
% TODO rephrase the above, it'll confuse people. Just need to reiterate the
% points about events, and maybe lead into the next section
Finally, gARGs are not defined in terms of the
coalescent with recombination (or any other) stochastic process.
Thus, the gARG encoding is focused on faithfully representing the
\emph{outcome} of the evolutionary processes of genetic inheritance
with recombination,
and is agnostic to the details of those processes.
% The paragraph above is a bit weak and muddled, needs a few more passes.
% In particular, the "Finally" bit doesn't really seem to follow
% from the others. Is there a more direct way we can address these points ?

@@

\section*{Event ARGs}\label{eARG}
Genome ARGs are composed by
describing the details of genetic inheritance
between specific ancestor and descendant genomes.
In contrast, ARGs are classically defined
by the evolutionary \emph{events} in the history of a sample.
The difference between gARGs and this prevalent view of
what defines an ARG is subtle and difficult to distinguish.
It is therefore useful to first survey some representative
recent definitions of
an ARG from the literature, to allow us to capture their common features
(see Appendix XXX for a full review).

\citet{brandt2021evaluation} follow the original Griffiths definition of
an ARG by coupling it directly to a generative process, stating
that the ``full ancestral recombination graph (ARG) is a structure that encodes all
coalescence and recombination events resulting from the stochastic process of
the coalescent with recombination.''
\citet{shipilina2023origin} emphasise the true existence of the
corresponding events (rather than the generative process) by saying that
``[t]he ARG describes the complete
ancestry of a sample of genomes through a series of real coalescence and
recombination events.''
\cite{mathieson2020ancestry}
define an ARG as ``a subset of the pedigree...
contain[ing] only those edges along which inherited segments of DNA have been
transmitted, and only those nodes corresponding to ancestors in which there was
a recombination or coalescence event.''
% In contrast, ~\citet{gusfield2014recombinatorics} defines an ARG as [TODO]
% Papers that do not discuss how positions work:
% - Mathieson and Scally: no discussion of how it works concretely (but quite
%   close to gARG, in some ways)
% TODO list out some of these as examples of where people don't bother
% getting into this important detail?

The common feature of these definitions is that they are expressed
in terms of evolutionary \emph{events}, and so for the purposes of
discussion we can label
this prevailing interpretation as an ``Event ARG'', or eARG.
(Many definitions also emphasise
the necessity of exhaustively capturing \emph{all} events, which we
examine in later sections.)
Another common feature of these definitions is that they do not explicitly
discuss the mechanism by which genealogical trees vary along the genome.
The topology of an eARG defines the common ancestor and
recombination events that occured,
but the graph structure alone is not sufficient to define the local trees,
or equivalently, how ``ancestral material'' (Section XXX) flows along
the edges of the ARG.
This is usually assumed to follow the mechanism described by
Griffiths and colleagues in which we have
two different ``types'' of node in the graph:
common ancestor nodes in which the inbound lineages are merged into a
single ancestral lineage with one parent, and recombination
nodes in which a single lineage is split into two independent
ancestral lineages.
Recombination nodes are then
annotated with
the corresponding crossover breakpoint~\citep{griffiths1996ancestral}.
Fig.~\ref{fig-arg-data-structure} shows an example of a classical
eARG with three samples (\noderef{a}, \noderef{b}, and \noderef{c})
and a single recombination event, along with an example concrete
encoding of the data.  There is a single recombination event
at node \noderef{d} with a breakpoint at position $x$. We
assume that \noderef{d} inherits genetic material to the
left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.
As shown in Fig.~\ref{fig-arg-data-structure}B,
the local trees are embedded in the graph, and can be reconstructed
in essentially the same manner as outlined the previous section for
gARGs. The only difference is that decisions about which outbound
edge to follow during the rootward traversal are taken at recombination
nodes, following
either the common convention that the order of parents
is significant~\citep[e.g.][]{griffiths1991two},
or by incorporating additional information to
distinguish left and right parents
into the
encoding~\citep[e.g.][]{gusfield2014recombinatorics,ignatieva2021kwarg}.

@@

\section*{Prospective vs retrospective ARGs}
Event ARGs, as defined in the previous section as a way of encompassing
the range of definitions provided in the literature, are
inherently \emph{retrospective}.
Event ARGs are defined in terms
of the events that occur in the history of some set of sampled
sequences.
For example, Fig.~\ref{fig-arg-in-pedigree}
is retrospective because it only includes information
about genetic inheritance relative to the sample individuals
$D_1$ and $D_2$ (genomes \textsf{a}--\textsf{d}).
That is, only the inheritance
history of the sample genomes is recorded in the ARG.
% Gregor & Jerome think that "inheritance" is clearer here than "transmission"
Genetic inheritance must have occurred between (e.g.)
$D_8$ and her daughter $D_6$, but this is not recorded because
it is not relevant to the genetic ancestry of the sample

@@

Prospective ARGs arise in individual-based forwards-time
simulations~\citep{kelleher2018efficient,haller2018tree},
where the genetic inheritance between genomes is recorded
exhaustively and periodically
``simplified'' to retain only
the ancestry relevant to the current population.
\citet{kelleher2018efficient} refer to the structure
recording all such forwards-in-time genetic relationships
as an ``embellished pedigree`` (and suggest the term ``nedigree'' as a shorthand).
Describing this instead as a ``prospective ARG'', emphasising
the use of the same data structures and the direct connection with
the usual retrospective interpretation of an ARG, may be
more useful.
The ``simplify'' algorithm, which prunes away parts of the graph that
are not ancestral to a sample, is the key enabling factor for
this approach to forwards-time simulation and is
described in detail by
\citet{kelleher2018efficient}.
It is a powerful tool for simulation and analysis, as it
allows to efficiently subset ARGs both in terms of the
samples and levels of detail about recombination retained,
as we discuss in the following sections.

\section*{Converting an eARG to a gARG}
Any eARG can be converted to a gARG without loss of information,
and the conversion process is very similar to
simplifying a prospective gARG with respect to a set of samples.
The process involves two steps, and is illustrated in
Fig.~\ref{fig-ancestry-resolution}.
The first step is straightforward
(Fig.~\ref{fig-ancestry-resolution}B).
We duplicate the nodes and
edges from the input eARG, and then add inheritance annotations
to the output gARG's edges. If the node is a common ancestor node,
we annotate the single outbound edge with the interval $[0,L)$,
for a genome of length $L$.
If the node is a recombination event with a breakpoint $x$,
we annotate the outbound edges with the intervals $[0, x)$
and $[x, L)$.
These inheritance annotations are clearly in one-to-one
correspondance with the information in the input eARG,
but they are a superset of the
``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
of the sample.

@@

\section*{ARGs and local trees}
Local trees, describing the genetic genealogy of a sample
at a particular location along the genome, are said to be
``embedded'' within in ARG [CITES], and the relationship
between the two is subtle and important.
% An ARG is sometimes referred to as being equivalent to a set of
% local trees along the genome [TODO citation], but the connection
% between the two is subtle.
% In order for the mapping to be one-to-one, so that the original
% gARG can be exactly recovered from a set of local trees and
% their genome coordinates,
% some specific properties of the ARG and the trees are required.
Note that many authors use the term ``marginal''
trees~\citep[e.g.][]{griffiths1996ancestral,minichiello2006mapping} [more]
which we avoid due the potential confusion with the
statistical concept of marginal distributions
(which could lead to interpretation as an integrated tree over many positions).

@@

The nodes in a local tree encoded
in this way are \emph{labelled} (i.e.\ carry the identity
of the corresponding ARG node), and this is fundamentally
important to how a local tree relates to its parent ARG.
If we are given the sequence of local trees for a gARG
encoded as an oriented tree (or in any other form in which
the tree nodes are explicitly labelled with the identity of the
corresponding ARG node)
along with the genome interval covered by each tree,
then we can recover precisely the same ARG. Thus, there
is a one-to-one correspondance between an ARG and
the sequence of local trees that it encodes.

The internal nodes of trees, however, are usually \emph{unlabelled},
because they most often correspond
to anonymous hypothetical ancestors. Internal
ARG nodes likewise correspond to anonymous hypothetical
ancestors, but if we do not label them we lose the fact that
the \emph{same} ancestors persist across multiple trees.
Without labelled internal tree nodes,
we must infer their identity
from the tree topology, which is both
expensive to compute and imprecise. For example
we could infer that node \textsf{k} is shared in the second
and third trees of in Fig.~\ref{fig-simplification}D
because it subtends identical subtrees in both,
but it would not be straightforward to infer that it
is also the root of the first tree.

\cite{shipilina2023origin} discuss the same ideas,
and note that
``full ARG ... contains more information than the series of tree
sequences along the genome''.
Their statement that an ARG
contains more information than its local trees is true only if
the local trees are unlabeled.
However, as we have shown, if the local trees
contain labelled internal (possibly unary) nodes,
then  there is precisely as
much information in them as an ARG.
Whether such unary nodes are \emph{observable}
is an entirely separate question, of course.

These properties have interesting consequences when we
consider the inference strategy of estimating local trees
independently and subsequently stitching them together
into an ARG.
Two recent methods take this approach,
\relate\ and
\espalier~\citep{rasmussen2022espalier},
and have different strategies for reconciling the local trees into an ARG.
\relate\ first infers local tree topologies, and then identifies
edges that persist from tree to tree by reasoning about the sets of samples
subtended. It does not attempt to be exhaustive.
\espalier, on the other hand, is specifically interested
in the details of recombination events (among small samples)
and therefore attempts to infer
the precise subtree prune-and-regraft (SPR)
operations~\citep{hein1990reconstructing,song2003on,song2006properties}
induced by recombination.
Inferring the SPRs between leaf-labelled trees is
NP-hard~\citep{hein1996complexity,allen2001subtree,bordewich2005computational},
but it is unclear what the complexity is when there
is a degree of internal node sharing between trees.

@@

\section*{Levels of simplification}\label{ARG_simplification}
ARG simplification is a powerful tool, and can do much more
besides removing unreachable nodes and edges in a prospective ARG
and resolving ancestral material in an eARG.
In general, we can think of
ARG simplification as the process
of removing nodes and re-writing edges (and their inheritance annotations)
to remove various types of redundancy.
The redundancy that we are interested in
revolves around the presence of nodes that appear
``unary'' in one or more local trees
(where a node has exactly one child over a particular genomic region).
These arise in several different ways.
Nodes with one child are not a standard feature of evolutionary trees;
we are usually interested in
nodes that have at least two children, because these
are in principle detectable from mutational differences
arising in the respective subtrees.
A tree branch may correspond to many generations
of individuals (or indeed cell divisions) and there is usually
very little information about these intermediates, or utility
in modelling them.
% FIXME the bridge between the next few paragraphs is a bit weird, in
% that we've just been talking about how interesting unary nodes are
% but then go on about how to remove them. Just needs a bit of smoothing
% out I think.
ARGs, however, are different, and there
can be significant information about evolutionary processes
provided by nodes that are unary in one or more local trees.
[FIXME this is clumsy; should mention somewhere, though.]
\citet{mathieson2020ancestry} show examples in which the ARGs
contain unary nodes, and state that paths are ``implicitly passing through points representing specific
individuals.''
% Full quote from fig caption:
% Genetic ancestry in the form of the ARG for a single individual. Combining
% genetic ancestry from different positions leads to a graph, incorporating all
% realized genetic ancestry paths, implicitly passing through points representing
% specific individuals.

@@

Simplifying away diamonds will remove many unary nodes from the
local trees, but there can still be nodes that are unary in all
of the local trees. In particular, a node can represent a recombinant
with multiple parents in the graph but only a single child (e.g.\ node \noderef{n}
in Fig.~\ref{fig-simplification}B), or can represent a common ancestor with
multiple children in the graph but in which no coalescence takes place
in the local trees
(node \noderef{r} in Fig.~\ref{fig-simplification}B).
% point out that this is why Hudson refers to CA nodes, not coalescence nodes
Such nodes are not singly connected in the graph, but are nevertheless unary in
all of the local trees.
The operation to remove them
therefore requires knowledge not just of the graph topology but also of the
ancestral material associated with the edges.
As we see in Fig.~\ref{fig-simplification}C,
removal of recombinant nodes can produce graph nodes with
more than two parents (e.g.~node \noderef{e}); and likewise, removal of
common ancestor but non-coalescent nodes can produce graph nodes with
more than two children (e.g.~node \noderef{s}). These represent a ``stacking up'' of
the \emph{effects} of multiple evolutionary events in a single node (genome), and the
ARG no longer contains the intermediate genomes representing those events.

@@

The remaining nodes are MRCAs of some subset of the samples
at \emph{some} positions along the genome. We still have
some unary nodes in the local trees, but these nodes will
correspond to a coalescence in at least one other
local tree. For example, node  \noderef{k} is unary in the fourth tree
of Fig.~\ref{fig-simplification}C, but is either binary
or ternary in all other local trees (recall this is a Wright-Fisher
simulation). The final level of simplification is to alter the edge annotations
such that, although no nodes are removed from the graph, all
unary nodes disappear from the local trees (Fig.~\ref{fig-simplification}D).
Note that although this last stage produces simpler local trees, by
removing information about the exact paths taken by lineages through
the graph, we lose potentially useful information about shared edges
between trees. The extent to which this information can be inferred
and utilised is an interesting open question.
% I don't think there's an actual manuscript or title to refer to here
% [CITE Ralph in prep.].
